---
title: "Logistic Regression -- The Challenger Disaster"
subtitle: "Bayesian / Frequentist Comparison"
author: "David Noonan"
date: "`r Sys.Date()`"
output:
  prettydoc::html_pretty:
    theme: architect
    highlight: github
---

The challenger disaster dataset is great for demonstrating logistic regression. Here I'm going to run frequentist and bayesian versions of a logistic regression model, and compare/contrast. Firstly, we need to load the dataset.  I found it conveniently in the DAAG package, so let's load it up:
```{r, message=FALSE, warning=FALSE, paged.print=FALSE}
#DAAG library contains the challenger dataset
library(DAAG)
#load the challenger "o-rings" dataset
chall <- orings
str(chall)
```

We have essentially four variables here:  "Temperature" -- the air temperature at "launch" (experimental launch simulation, I believe), "Blowby" -- when the o-ring leaks fuel, and "Erosion" -- where the o-ring erodes, and "Total" -- a combination of "Erosion" and "Blowby".  We are particularly interested in temperature, which we'll use to explain the outcome, and we are also interested in "Total", which combines both types of o-ring failures that cause the shuttle to blow up. Below is a chart of "Total" by temperature:

```{r, fig.margin = TRUE}
#plot failure by temperature
library(ggplot2)
ggplot() +
  geom_col(data = chall,
           mapping = aes(x = Temperature,
                         y = Total)
  ) +
  ggtitle(label = "O-ring Failure by Temperature") +
  xlab("Temperature (f)") +
  ylab("Failure")
```
 
 
We can already see what looks like a trend:  Lower temperatures cause more failures. Since we're not interested in the number of o-ring failures, or the type, we create a new variable, "Failure", which indicates the presence or absence of failure: 

```{r}
library(tidyverse)
#Create failure variable, 1 = failure, 0 = no failure
challenger <- orings %>% mutate(failure = ifelse(test = Total > 0,
                                            yes =  1,
                                            no =  0))
```

